---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# README

<!-- badges: start -->
<!-- badges: end -->

In this R project, all the scripts for the manuscript *Plankton blooms over the annual cycle shape trophic interactions under climate change* are available.

All the analyses were done with `r R.version[c("version.string")]` on the platform `r R.version[c("platform")]`.

Data and scripts are also stored on Zenodo https://zenodo.org/records/10708263


# R scripts

```{r, echo=TRUE, eval=FALSE}
# 1. Download the data
source("Scripts/DownloadData.R")

# 2. Calculate the phenology metrics and the phyisical variables
source("Scripts/PhenologyZP.R")
source("Scripts/PhenologyPP.R")
source("Scripts/Physical.R")

# 3. Run the analysis
source("Scripts/PlanktonSuccession.R")
source("Scripts/Trends.R")
source("Scripts/GLM.R")

# 4. Run the supplementary analyses
source("Scripts/DifferenceStations.R")
source("Scripts/Nauplii.R")
source("Scripts/FreqPhytoProp.R")
source("Scripts/Validation_Trends.R")
```

### DownloadData.R {#customscript0}

This script can be used to download the phytoplankton and zooplankton time series data as well as the abiotic parameter.


### PhenologyZP.R {#customscript1}

In this script the *Zooplankton metrics* are calculated:

* Imported raw abundance data from [Sharkweb](https://sharkweb.smhi.se/hamta-data/)[^1]
* Biomass was calculated from individual bodymass based on location, life stage and sex according to the [HELCOM COMBINE manual](https://helcom.fi/action-areas/monitoring-and-assessment/monitoring-guidelines/combine-manual/)
* Species were filtered to keep only the one with a good data coverage and excluding Nauplii
* Density plot were performed based on the daily interpolation with a kernel smoothing
* Weekly interpolation were made over the full time series with the Rpackage *zoo* and Week were retrieved using the package *lubridate*
* Metric were calculated
  + Peak timing using the center of gravity method
  + Duration of the bloom using the cumulative biomass between the 25^th^ and 75^th^ percentile
  + Magnitude of the bloom using the average of the biomass between the start and the end of the bloom
  + A final dataset was saved combining all metric together **"Output/Data/Weekly_Full_ZP.txt"**

### PhenologyPP.R {#customscript2}

In this script the *Phytoplankton metrics* are calculated:

* Imported raw biomass data from [Sharkweb](https://sharkweb.smhi.se/hamta-data/)[^1]
* Species were filtered to keep only the ones with a good data coverage and exclude heterotroph organisms
* Density plot were performed based on the daily interpolation with a kernel smoothing
* Spring bloom was the sum of the phytoplankton between Week 5 and 2
* Summer bloom was the sum of the phytoplankton between Week 22 and 37
* Weekly interpolation were made over the full time series with the Rpackage *zoo* and DOY were retrieved using the package *lubridate*
* Metric were calculated
  + Peak timing using the center of gravity method
  + Duration of the bloom using the cumulative biomass between the 25^th^ and 75^th^ percentile
  + Magnitude of the bloom using the average of the biomass between the start and the end of the bloom
* A final dataset was saved combining all metric together **"Output/Data/Weekly_Full_PP.txt"**

### Physical.R {#customscript3}

In this script the *Abiotic factor* are calculated:

* Imported raw data from [Sharkweb](https://sharkweb.smhi.se/hamta-data/)[^1]
* Weekly interpolation were made over the full time series with the Rpackage *zoo* and DOY were retrieved using the package *lubridate*
* Monthly average are saved as **"Output/Data/Abiotic.txt"**

[^1]: Note that data were downloaded using the [shaRk](https://github.com/andreasnovotny/shaRk) Rpackage for this analysis. If the data are downloaded throughout the [Sharkweb](https://sharkweb.smhi.se/hamta-data/), make sure to update the headers accordingly.

### PlanktonSuccession.R

In this script the **Fig. 1b** and **Fig. 3** are made:

* Imported metrics and abiotic factor from 
  + **"Output/Data/Weekly_Full_ZP.txt"**
  + **"Output/Data/Weekly_Full_PP.txt"**
  + **"Output/Data/Abiotic.txt"**
* Panel a shows the water temperature between 0 to 20m for each month at each station (average and range min-max)
* Panel b shows the salinity between 0 to 20m for each month at each station (average and range min-max)
* Panels c and d show total and taxa specific dynamics
  + The peak timing (average between 2008 and 2022)
  + The start and end of the bloom (average between 2008 and 2022)
* A map is also plotted at the end of the script

### Trends.R

In this script the *Trends* are analysed:

* Imported data from 
  + **"Output/Data/Weekly_Full_ZP.txt"**
  + **"Output/Data/Weekly_Full_PP.txt"**
  + **"Output/Data/Abiotic.txt"**
* Trends are analysed with the Kendall test from the *Kendall* package and effect size was assessed using the Sen's slope from the *mblm* package for phytoplankton, zooplankton and seasonal salinity and temperature:
  + Peak timing
  + Peak initation
  + Peak termination
  + Peak magnitude
* Coefficients of the trends for the phenological metrics are shown in **Fig. 4** and **Fig. S3**
* z-scores of the peak timing and magnitudes were calculated for all taxa and visualized across the year in **Fig. 4** and **Fig. S3**
* Coefficients of the trends and dynamics across the years for the abiotic parameters are shown in **Fig. 2** and **Fig. S2**

### GLM.R {#customscript4}

In this script the Generalized Linear Models are made:

* Imported data from 
  + **"Output/Data/Weekly_Full_ZP.txt"**
  + **"Output/Data/Weekly_Full_PP.txt"**
  + **"Output/Data/Abiotic.txt"**
* Output of the GLM models are shown in **Fig. 5** and **Table S4**


### Nauplii.R

In this script **Fig. S5** is produced showing seasonal abundance dynamics of *Acartia*, *Pseudocalanus*, *Temora* and *Centropages* nauplii stages are plotted 

### FreqPhytoProp.R

In this script the **Fig. S1** and **Fig. 1** are produced:

* For all three stations, based on:
  + Zooplankton sampling
  + Phytoplankton sampling
  + Temperature sampling
* Density plot showing the total phytoplankton and zooplankton biomass dynamics and the contribution of each taxa  

### Validation_Trends.R

In this script **Fig. S6** is produced:

* We subsampled phytoplankton and zooplankton to have a similar temporal resolution than at BY15 and BY5
* We weekly interpolated the data and calculated the phenology metrics for all taxa at BY31
* We compared the subsampled data with the complete dataset

---

# Data

First, the **raw data** were downloaded and are publicly available:

  - Manually by going on [Sharkweb](https://sharkweb.smhi.se/hamta-data/)
  - Using the shaRk package and the script `DownloadData`

Note that data were downloaded using the [shaRk](https://github.com/andreasnovotny/shaRk) Rpackage for this analysis. If the data are downloaded throughout the [Sharkweb](https://sharkweb.smhi.se/hamta-data/), make sure to update the headers accordingly.

  
The zooplankton bodymass were retrived from the [HELCOM COMBINE manual](https://helcom.fi/action-areas/monitoring-and-assessment/monitoring-guidelines/combine-manual/), are shown in the Table S1 of the manuscript, and should have this structure to be compatible with the scripts:
```{r, echo=FALSE}
read.csv("Data/Bodymass.csv") |> str()
```

Where:

* `Taxa` corresponds to the Taxa of interest
* `STAGE` corresponds to the Stage of the organism (`AD`, `C1`, `C4`, `JV` or `NS`)
* `SEXCO` corresponds to the sex code of the organism (`F`, `M`, `NS`)
* `Station`corresponds to the station (`BY31`, `BY15` or `BY5`)
* `Jan.Mar` corresponds to the bodymass in Winter
* `Apr.Jun` corresponds to the bodymass in Spring
* `Jul.Sep` corresponds to the bodymass in Summer
* `Oct.Dec` corresponds to the bodymass in Autumn

## Processed data

In the folder **Output/Data** all the needed data can be found.

1. Weekly_Full_ZP.txt
2. Weekly_Full_PP.txt
3. Abiotic.txt

## Weekly_Full_ZP.txt {#customdata1}

This dataset is produced at the end of the script `Phenology_ZP.R`

It has this structure:

```{r, echo=FALSE}
read.delim("Output/Data/Weekly_Full_ZP.txt", sep = ";") |> str()
```

Where:

* `Taxa`: Mesozooplankton taxa: *character*
* `Year`: Year between 2008 and 2022, *integer*
* `Station`: the three stations BY31, BY15, and BY5, *character*
* `Group`: Group of mesozooplankton taxa: Copepoda, Cladocera, and Rotatoria, *character*
* `Timing`: Bloom peak timing (Week #), *numeric*
* `Maximum`: Bloom magnitude maximum ($\mu$g L^-1^), *numeric*
* `Magnitude`: Bloom magnitude ($\mu$g L^-1^), *numeric*
* `Start`: Start of the growing season, correspond to the 25^th^ percentile of the cumulative biomass (Week #), *numeric*
* `End`: End of the growing season, correspond to the 75^th^ percentile of the cumulative biomass (Week #), *numeric*
* `Middle`: Middle of the growing season, correspond to the 50^th^ percentile of the cumulative biomass (Week #), *numeric*
* `Duration`: Time span between the Start and the End of the growing season (Week #), *numeric*



## Weekly_Full_PP.txt {#customdata2}

This dataset is produced at the end of the script `Phenology_PP.R`

It has this structure:

```{r, echo=FALSE}
read.delim("Output/Data/Weekly_Full_PP.txt", sep = ";") |> str()
```

Where:

* `Taxa`: Phytoplankton taxa, *See Table S3 for more information* *character*
* `Year`: Year between 2008 and 2021, *integer*
* `Station`: the three stations BY31, BY15, and BY5, *character*
* `Group`: Group of phytoplankton taxa: only Phytoplankton, *character*
* `Timing`: Bloom peak timing (Week #), *numeric*
* `Maximum`: Bloom magnitude maximum ($\mu$gC L^-1^), *numeric*
* `Magnitude`: Bloom magnitude ($\mu$gC L^-1^), *numeric*
* `Start`: Start of the growing season, correspond to the 25^th^ percentile of the cumulative biomass (Week #), *numeric*
* `End`: End of the growing season, correspond to the 75^th^ percentile of the cumulative biomass (Week #), *numeric*
* `Middle`: Middle of the growing season, correspond to the 50^th^ percentile of the cumulative biomass (Week #), *numeric*
* `Duration`: Time span between the Start and the End of the growing season (Week #), *numeric*

## Abiotic.txt {#customdata3}

This dataset is produced with the script `Physical.R`

It has this structure:

```{r, echo=FALSE}
read.delim("Output/Data/Abiotic.txt", sep = ";") |> str()
```

Where:

* `Year`: Year of the sampling event, *integer*
* `month`: Month of the sampling event. 1 corresponds to January, 2 to February, and so on, *integer*
* `Parameter`: Parameter sampled. *character*
  + Oxy_0.20: Integrated oxygen concentration between 0 and 20m depth, mL L^-1^
  + SST: Sea surface temperature, °C
  + Temp_0.20: Integrated temperature between 0 and 20m depth, °C
  + Salinity: Integrated salinity between 0 and 20m depth
  + Salinity_SSS: Sea surface salinity
  + Nutrient: Nitrate and nitrite concentration between 0 and 20m depth, $\mu$mol L^-1^
  + strat_index: Stratification index calculated as (SST - Temperature at 60m) / average temperature between 0 and 60m
* `Station`: Station sampled, *character*
* `Value`: value of the parameter, *numeric*
* `n`: number of observation, *integer*